Testing Statistical Graphics

### 2022-06-29 index.qmd
Accept Emily's changes

# Testing Methods and \ear{User} Engagement 
<!-- Lineups, You Draw It, Direct Assessment -->

@@

Statistical lineups\cite{bujaStatisticalInferenceExploratory2009,wickhamGraphicalInferenceInfovis2010} are a more generally applicable testing tool for perceptual questions such as "which chart displays this data more clearly"\cite{power}. 
As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot instead of any of the 19 null plots, then the target plot is said to be visually significant\cite{majumderValidationVisualStatistical2013,vanderplasStatisticalSignificanceCalculations2021} and a "see" value (the visual analogue of a p-value for a statistical test) can be calculated\cite{niladriroychowdhurySeeValueApp2020}.
In another variation of the statistical lineup procedure, two models are compared, with target plots from each model embedded in the array of 20 total plots; the remaining null plots are constructed from a mixture model that blends the two competing models. This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data \ear{and assign a hiearchy of visual attributes}\cite{clusters}.
Statistical lineups do not require the experimenter to provide any contextual information for the data: all of the necessary information is embedded in the choice of a null statistical model, which is convenient for testing purposes, but does not require engagement at a level that would allow for testing the viewer's understanding of the information shown in the chart. 
An example of a statistical lineup for testing the perception of exponential growth is shown in @fig-lineup-log.

@@

![This question asks users to write out a description of how the population of Ewoks changes over time, without any further cues, to determine whether participants default to multiplicative or additive \ear{language} descriptions.](Estimation_describe_plot_crop.png){#fig-estimation-describe}

@@

![This question asks participants for a numerical estimate, but provides a basic calculator and scratchpad. All user interactions with the calculator and scratchpad are logged, providing insight into the user's thought process \ear{and estimation strategy}.](Estimation_numerical_screenshot_crop.png){#fig-estimation-calc}

In our experience, one of the most difficult components of designing an experiment which asks users to directly estimate information from a chart using a full scenario (background information, etc. as well as contextual details from the chart) is that the \ear{selection and} wording of such questions must be extremely carefully determined.
\ear{Mathematics education researchers have provided guidelines for selecting different levels of questioning in order to assess graph comprehension: literal reading of the data, reading between the data, and reading beyond the data \cite{wood1968objectives, curcio1987comprehension, friel2001making, glazer2011challenges}. In a recent study, we identified questions based on this framework for evaluating direct estimates and extending those estimates to make comparisons between two points.}
<!-- , and e -->
\ear{E}ven when great care is taken with the construction of the question, participant answer accuracy is fundamentally limited by the fact that many participants do not read and interpret the question with the care and precision that it was written.
Questions that ask participants to e.g. estimate the multiplicative \ear{change} in a quantity at two time points may be misunderstood as asking for an estimate of the additive difference, and the resulting estimates are then one or more orders of magnitude off of the correct answer.
This is one area where lineup tests are convenient - they do not depend on participants to understand the nuances of language or scenarios built around the chart under investigation.
However, in some situations it may be sufficient to ask participants to estimate direct numerical quantities that have little contextual information, as we did in \cite{vanderplasFramedReproducingRevisiting2019} when assessing the accuracy of framed plots re-created from the Statistical Atlas.

It is also possible to design a task that requires participants to *engage directly* with an interactive visualization. 
This can be useful in a directed task, where users are asked to interact with the chart in a specific way and the result is recorded, but it is also possible to use interactive visualizations in an open-ended task, recording how users engage with the graphic in an exploratory (as opposed to goal-directed) manner. 
In one recent experiment, we asked participants to forecast an exponential trend, with data presented on either a linear or log scale. 
Using JavaScript code modified from New York Times interactive graphics "You Draw It" features\cite{katzYouDrawIt2017}, we had users draw trend lines \ear{with their computer mouse} and make forecasts directly on interactive charts, with the data and user-drawn predictions recorded to our database.
With interactive graphics rendered using JavaScript (or other web libraries), the only limit to the types of questions you can ask in testing graphics is in your ability to write code to interact with the visualization library. 
This type of testing method can be extremely natural for participants, but it also is hard to generalize when discussing testing methods because of the potential range of applications where it might be employed.

Whichever testing method is chosen should be appropriate to the type of question under investigation and the level of visual \ear{and cognitive} engagement required to answer that question. While lineups are excellent tools for assessing perceptual questions, they cannot address questions aimed at understanding how people use charts within the wider context of a story or practical task; this requires more direct methods with higher ecological validity.

\ear{kind of a sudden jump into this paragraph, but I'm not sure how to transition into it right now.}
Lineups require a great amount of work in developing a null sampling model which replicates the important visual features of the data, but model specification is extremely important when asking users questions about any simulated data set or even when examining user interactions directly. 
It  is important to carefully calibrate the model parameters to allow for some variability, but not too much, and to ensure that participants can succeed at the task and do not feel like they are being made to analyze random noise. 
This Goldilocks-style problem is the focus of the next section.

@@

The primary human design factor to be aware of is that visual tasks and assessing statistical graphics can be extremely cognitively taxing. In our experience, it is difficult to expect participants to evaluate more than about 15 statistical lineups in one sitting. It is easy to understand how this arises when we consider the concept of decision fatigue: each standard lineup has 20 panels, so there are 190 pairwise comparisons to evaluate between the plots of just a single lineup (though it is unlikely that participants actually sequentially evaluate each comparison in a combinatorial fashion). Tasks which are more interactive, such as \ear{`You Draw It'}, may be somewhat easier for participants, but it is unlikely that participants would be willing to complete more than about 20 tasks in one sitting even with tasks that require fewer decisions and more engagement. At some point, participants' interest in accuracy will decline, and the researcher is better off fitting the task within the window where participants are attentive and engaged.

It can be extremely helpful to include "practice" demonstrations of the task to show the basic process, logic, and reasoning. While it is tempting to make these tasks fully representative of the type of judgement which will be required of participants, practice tasks which are too close to the experimental task may bias participants; we have found that it works well to have a relatively easy practice task which utilizes a slightly different type of plot and/or type of data than what will be tested in the experiment. In cases which require interactivity, gif animations of the task being carried out are useful, as are additional visual cues, such as the yellow box used in the \ear{`You Draw It'} task[^1] to indicate that there were points which were not completed.

>>>>

# Testing Methods and User Engagement 
<!-- Lineups, You Draw It, Direct Assessment -->

@@

Statistical lineups\cite{bujaStatisticalInferenceExploratory2009,wickhamGraphicalInferenceInfovis2010} are a more generally applicable testing tool for perceptual questions such as "which chart displays this data more clearly"\cite{power}. 
As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot instead of any of the 19 null plots, then the target plot is said to be visually significant\cite{majumderValidationVisualStatistical2013,vanderplasStatisticalSignificanceCalculations2021} and a "see" value (the visual analogue of a p-value for a statistical test) can be calculated\cite{niladriroychowdhurySeeValueApp2020}.
In another variation of the statistical lineup procedure, two models are compared, with target plots from each model embedded in the array of 20 total plots; the remaining null plots are constructed from a mixture model that blends the two competing models. This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data and assign a hiearchy of visual attributes\cite{clusters}.
Statistical lineups do not require the experimenter to provide any contextual information for the data: all of the necessary information is embedded in the choice of a null statistical model, which is convenient for testing purposes, but does not require engagement at a level that would allow for testing the viewer's understanding of the information shown in the chart. 
An example of a statistical lineup for testing the perception of exponential growth is shown in @fig-lineup-log.

@@

![This question asks users to write out a description of how the population of Ewoks changes over time, without any further cues, to determine whether participants default to multiplicative or additive language descriptions.](Estimation_describe_plot_crop.png){#fig-estimation-describe}

@@

![This question asks participants for a numerical estimate, but provides a basic calculator and scratchpad. All user interactions with the calculator and scratchpad are logged, providing insight into the user's thought process and estimation strategy.](Estimation_numerical_screenshot_crop.png){#fig-estimation-calc}

In our experience, one of the most difficult components of designing an experiment which asks users to directly estimate information from a chart using a full scenario (background information, etc. as well as contextual details from the chart) is that the selection and wording of such questions must be extremely carefully determined.
Mathematics education researchers have provided guidelines for selecting different levels of questioning in order to assess graph comprehension: literal reading of the data, reading between the data, and reading beyond the data \cite{wood1968objectives, curcio1987comprehension, friel2001making, glazer2011challenges}. In a recent study, we identified questions based on this framework for evaluating direct estimates and extending those estimates to make comparisons between two points.

Even when great care is taken with the construction of the question, participant answer accuracy is fundamentally limited by the fact that many participants do not read and interpret the question with the care and precision that it was written.
Questions that ask participants to e.g. estimate the multiplicative change in a quantity at two time points may be misunderstood as asking for an estimate of the additive difference, and the resulting estimates are then one or more orders of magnitude off of the correct answer.
This is one area where lineup tests are convenient - they do not depend on participants to understand the nuances of language or scenarios built around the chart under investigation.
However, in some situations it may be sufficient to ask participants to estimate direct numerical quantities that have little contextual information, as we did in \cite{vanderplasFramedReproducingRevisiting2019} when assessing the accuracy of framed plots re-created from the Statistical Atlas.

It is also possible to design a task that requires participants to *engage directly* with an interactive visualization. 
This can be useful in a directed task, where users are asked to interact with the chart in a specific way and the result is recorded, but it is also possible to use interactive visualizations in an open-ended task, recording how users engage with the graphic in an exploratory (as opposed to goal-directed) manner. 
In one recent experiment, we asked participants to forecast an exponential trend, with data presented on either a linear or log scale. 
Using JavaScript code modified from New York Times interactive graphics "You Draw It" features\cite{katzYouDrawIt2017}, we had users draw trend lines with their computer mouse and make forecasts directly on interactive charts, with the data and user-drawn predictions recorded to our database.
With interactive graphics rendered using JavaScript (or other web libraries), the only limit to the types of questions you can ask in testing graphics is in your ability to write code to interact with the visualization library. 
This type of testing method can be extremely natural for participants, but it also is hard to generalize when discussing testing methods because of the potential range of applications where it might be employed.

Whichever testing method is chosen should be appropriate to the type of question under investigation and the level of visual and cognitive engagement required to answer that question. While lineups are excellent tools for assessing perceptual questions, they cannot address questions aimed at understanding how people use charts within the wider context of a story or practical task; this requires more direct methods with higher ecological validity.

<!-- \ear{kind of a sudden jump into this paragraph, but I'm not sure how to transition into it right now.} -->
Lineups require a great amount of work in developing a null sampling model which replicates the important visual features of the data, but model specification is extremely important when asking users questions about any simulated data set or even when examining user interactions directly. 
It  is important to carefully calibrate the model parameters to allow for some variability, but not too much, and to ensure that participants can succeed at the task and do not feel like they are being made to analyze random noise. 
This Goldilocks-style problem is the focus of the next section.

@@

The primary human design factor to be aware of is that visual tasks and assessing statistical graphics can be extremely cognitively taxing. In our experience, it is difficult to expect participants to evaluate more than about 15 statistical lineups in one sitting. It is easy to understand how this arises when we consider the concept of decision fatigue: each standard lineup has 20 panels, so there are 190 pairwise comparisons to evaluate between the plots of just a single lineup (though it is unlikely that participants actually sequentially evaluate each comparison in a combinatorial fashion). Tasks which are more interactive, such as `You Draw It', may be somewhat easier for participants, but it is unlikely that participants would be willing to complete more than about 20 tasks in one sitting even with tasks that require fewer decisions and more engagement. At some point, participants' interest in accuracy will decline, and the researcher is better off fitting the task within the window where participants are attentive and engaged.

It can be extremely helpful to include "practice" demonstrations of the task to show the basic process, logic, and reasoning. While it is tempting to make these tasks fully representative of the type of judgement which will be required of participants, practice tasks which are too close to the experimental task may bias participants; we have found that it works well to have a relatively easy practice task which utilizes a slightly different type of plot and/or type of data than what will be tested in the experiment. In cases which require interactivity, gif animations of the task being carried out are useful, as are additional visual cues, such as the yellow box used in the `You Draw It' task[^1] to indicate that there were points which were not completed.
